Live Functional Programming with Typed Holes

### 2018-04-12 calculus.tex
a bit more progress through Section 3; more to come later today

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% <<<<<<<<<<<<< Matt is here (latest pass over everything above is done -- 2018.04.11)

The rules governing expansion of constants, variables and lambda expressions --- Rules {ESConst}, {ESVar}, {ESLam} and {EALam} --- and the corresponding type assignment rules --- Rules {TAConst}, {TAVar} and {TALam} --- mirror the typing rules from Fig.~\ref{fig:bidirectional-typing} 
(so the corresponding cases of Theorem~\ref{thm:typed-expansion}, Theorem~\ref{thm:expandability} and Theorem~\ref{thm:expansion-generality} are straightforward). 
Note that in the internal language, all lambdas are half-annotated, again to support type assignment---Rule {EALam} inserts the annotation when expanding an unannotated external lambda based on the given type. 
The rules governing holes and the rules governing function applications and ascriptions, which both perform \emph{cast insertion}, are more interesting, so let us consider both cases in turn.

\subsubsection{Hole Expansion}\label{sec:hole-expansion} 
Rules {ESEHole}, {ESNEHole}, {EAEHole} and {EANEHole} govern the expansion of empty and non-empty expression holes to empty and non-empty \emph{hole closures}, $\dehole{u}{\sigma}{}$ and $\dhole{\dexp}{u}{\sigma}{}$. 
%

The hole name, $u$, on a hole closure identifies the external hole that the hole closure corresponds to. 
Note that while we assume each hole name to be unique in the external language, there can be multiple hole closures with the same name during evaluation due to substitution. 
For example, the result from Fig.~\ref{fig:grades-example} showed four closures for the hole named 1\todo{update with actual number (four?) and hole name later}. 
There, we numbered each hole closure for a given hole sequentially, \li{1:1}, \li{1:2} and so on, but this is strictly for the sake of presentation so we omit hole closure numbers from the core calculus.

The hole expansion rules are the only rules that introduce hypotheses, of the form $\Dbinding{u}{\hGamma}{\htau}$, into the hole context, $\Delta$. 
The purpose of the hole context is to record a type, $\tau$, and a typing context, $\Gamma$, for each hole name, $u$.\footnote{
We use a hole context, rather than recording the typing context and type directly on each hole closure, to ensure that all closures for a hole name have the same typing context and type.} 
This notation for hole contexts is taken from contextual modal type theory (CMTT) \cite{Nanevski2008}, identifying hole names with metavariables and hole contexts with modal contexts (we say more about the connection with CMTT below). 
In all four hole expansion rules, the typing context recorded in the hole context is simply the current typing context when the hole is expanded. 
In the synthetic hole expansion rules, {ESEHole} and {ESNEHole}, the generated hole context assigns the hole type, $\tehole$, to $u$, as in the typing rules. 
However, the first two premises of the expansion subsumption rule, Rule EASubsume, disallow the use of subsumption for holes in analytic position. 
Instead, we have separate analytic rules, {EAEHole} and {EANEHole}, which record the type that the hole is being checked against into the hole context. 
This is again so that we can use type assignment for the internal language --- the type assignment rules TAEHole and TANEHole in Fig.~\ref{fig:hasType} assign a hole closure for $u$ the corresponding type from the hole context.

Each hole closure also has an associated environment, $\sigma$, which is a finite substitution of the form $[d_1/x_1, ~\cdots, d_n/x_n]$ for $n \geq 0$. 
The purpose of the closure environment is to keep a record of the substitutions that occur around the hole as evaluation occurs. 
Initially, no evaluation has occurred, so the initial environment generated by the hole expansion rules is the identity substitution for the typing context associated with $u$ in $\Delta$, which we notate $\idof{\hGamma}$ and define as follows.
\begin{defn}[Identity Substitution] $\idof{x_1 : \tau_1, ~\cdots, x_n : \tau_n} = [x_1/x_1, ~\cdots, x_n/x_n]$
\end{defn}
\noindent
The type assignment rules for hole closures, Rules TAEHole and TANEHole, require that we be able to check the environment of each hole closure against the corresponding typing context, written $\hasType{\Delta}{\hGamma}{\sigma}{\hGamma'}$ and defined as follows:\todo{check that the definition in the Agda corresponds}
\begin{defn}[Substitution Typing]
$\hasType{\Delta}{\hGamma}{\sigma}{\hGamma'}$ iff $\domof{\sigma} = \domof{\hGamma'}$ and for each $x : \htau \in \hGamma'$ we have that $d/x \in \sigma$ and $\hasType{\Delta}{\hGamma}{d}{\tau}$.
\end{defn}
\noindent
It is easy to verify that the identity substitution satisfies this requirement, i.e. that $\hasType{\Delta}{\hGamma}{\idof{\hGamma}}{\hGamma}$.

Empty hole closures, $\dehole{u}{\sigma}{}$,  correspond to the metavariable closures (a.k.a. deferred substitutions) from CMTT, $\cmttclo{u}{\sigma}$.\todo{change notation from CMTT}{} 
We will see how closure environments evolve during evaluation in Sec.~\ref{sec:evaluation}. 
Non-empty hole closures, $\dhole{d}{u}{\sigma}{}$, do not directly correspond to a notion from CMTT (see Sec.~\ref{sec:resumption}).

>>>>

The rules governing expansion of constants, variables and lambda expressions
--- Rules {ESConst}, {ESVar}, {ESLam} and {EALam}
--- and the corresponding type assignment rules
--- Rules {TAConst}, {TAVar} and {TALam}
--- mirror the typing rules from Fig.~\ref{fig:bidirectional-typing}.
%
Consequently, the corresponding cases of
Theorem~\ref{thm:typed-expansion}, Theorem~\ref{thm:expandability} and
Theorem~\ref{thm:expansion-generality} are straightforward.
%
To support type assignment, all lambdas in the internal language are
half-annotated---Rule {EALam} inserts the annotation when expanding an
unannotated external lambda based on the given type.
%
Some rules perform \emph{cast insertion}, making them more interesting
than the others; these rules include those for holes, and the those for
function applications and ascriptions.  We consider each below.

\subsubsection{Hole Expansion}\label{sec:hole-expansion} 
Rules {ESEHole}, {ESNEHole}, {EAEHole} and {EANEHole} govern the expansion of empty and non-empty expression holes to empty and non-empty \emph{hole closures}, $\dehole{u}{\sigma}{}$ and $\dhole{\dexp}{u}{\sigma}{}$, respectively. 
%
The hole name~$u$ on a hole closure identifies the external hole to which the hole closure corresponds.
%
%Note that
While we assume each hole name to be unique in the external language, due to substitution, evaluated (internal) terms may consist of multiple hole closures with the same name.
%
For example, the result from Fig.~\ref{fig:grades-example} shows four closures for the hole named 1\todo{update with actual number (four?) and hole name later}.
%
There, we number each hole closure for a given hole sequentially, \li{1:1}, \li{1:2} and so on, but this is strictly for the sake of presentation, so we omit hole closure numbers from the core calculus.

The hole expansion rules introduce hypotheses of the form~$\Dbinding{u}{\hGamma}{\htau}$ into the hole context~$\Delta$.
%
For each hole name~$u$, the hole context records a type~$\tau$, and a typing context~$\Gamma$.\footnote{
We use a hole context, rather than recording the typing context and type directly on each hole closure, to ensure that all closures for a hole name have the same typing context and type.}
%
We borrow this hole context notation from contextual modal type theory (CMTT) \cite{Nanevski2008}, identifying hole names with metavariables and hole contexts with modal contexts (we say more about the connection with CMTT below).
%
Each hole expansion rule records the ``current'' typing context under which the hole is expanded.
%
In the synthetic hole expansion rules~{ESEHole} and~{ESNEHole}, the generated hole context assigns the hole type~$\tehole$ to hole name~$u$, as in the typing rules.
%
However, the first two premises of the expansion subsumption rule~{EASubsume} disallow the use of subsumption for holes in analytic position.
%
Instead, we employ separate analytic rules~{EAEHole} and~{EANEHole}, which each record the checked type~$\tau$ in the hole context.
%
Consequently, we can use type assignment for the internal language --- the type assignment rules TAEHole and TANEHole in Fig.~\ref{fig:hasType} assign a hole closure for hole name~$u$ the corresponding type from the hole context.

Each hole closure also has an associated environment~$\sigma$ which consists of a finite substitution of the form $[d_1/x_1, ~\cdots, d_n/x_n]$ for $n \geq 0$.
%
The closure environment keeps a record of the substitutions that occur around the hole as evaluation occurs.
%
Initially, when no evaluation has yet occurred, the hole expansion rules generate an initial environment that consists of the identity substitution for the typing context associated with hole name~$u$ in hole context~$\Delta$, which we notate $\idof{\hGamma}$, and define as follows.
%
\begin{defn}[Identity Substitution] $\idof{x_1 : \tau_1, ~\cdots, x_n : \tau_n} = [x_1/x_1, ~\cdots, x_n/x_n]$
\end{defn}
\noindent
The type assignment rules for hole closures~{TAEHole} and {TANEHole} each require that the hole closure environment~$\sigma$ be consistent with the corresponding typing context, written as $\hasType{\Delta}{\hGamma}{\sigma}{\hGamma'}$.
%
Formally, we define this relation as follows:\todo{check that the definition in the Agda corresponds}
\begin{defn}[Substitution Typing]
$\hasType{\Delta}{\hGamma}{\sigma}{\hGamma'}$ iff $\domof{\sigma} = \domof{\hGamma'}$ and for each $x : \htau \in \hGamma'$ we have that $d/x \in \sigma$ and $\hasType{\Delta}{\hGamma}{d}{\tau}$.
\end{defn}
\noindent
It is easy to verify that the identity substitution satisfies this requirement, i.e. that $\hasType{\Delta}{\hGamma}{\idof{\hGamma}}{\hGamma}$.

Empty hole closures, $\dehole{u}{\sigma}{}$,  correspond to the metavariable closures (a.k.a. deferred substitutions) from CMTT, $\cmttclo{u}{\sigma}$.\todo{change notation from CMTT}{}
%
\Secref{sec:evaluation} defines how these closure environments evolve during evaluation.
%
Non-empty hole closures~$\dhole{d}{u}{\sigma}{}$ have no direct correspondance with a notion from CMTT (see Sec.~\ref{sec:resumption}).

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% Matt is here (2018.04.12, ~8am MT)
